Application of Generalized Additive Models for long-term and seasonal trend analysis of water quality: South San Francisco Bay case study

### 2020-08-04 manu_draft.Rmd
gam0 removed from manu, added doc links to readme

\begin{equation}
\boldsymbol{gam0:}\ g\left(chl\right) = s_0 + \beta_1 year + s_1\left(doy\right) + \epsilon 
(\#eq:gamzero)
\end{equation}

@@

A challenge for optimizing GAMs to time series data with significant intra- and interannual variation is choosing an appropriate functional form (i.e., gam0, gam1, gam2, or gam6) and determining how much variation could be explained by the smoothers in each function.  The four models above provide tradeoffs in the functional forms that balance descriptive and computational efficiency (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).  However, the individual smoothing functions (i.e, $s_1()$, $s_2()$, $s_2^*()$ and $ti_1()$) require user input on the potential upper limit for the *k*-values that could be used during model fitting.  Initial testing by Murphy et al. (2019) showed that the default *k* values in the mgcv package were insufficient for describing the interannual variation in the response variable.  Instead, the upper limit for the number of knots for the $s_2(year)$ smoother was specified as the maximum between 10 or  $2 / 3 \times$ the number of years in the time series.   We follow the same approach herein for gam1 and gam2.  Further, the default *k*-values for $s_2()$ and $ti_1()$ were considered appropriate and not tested further. For the application to SFE chlorophyll data, gam6 was also added as an extension of gam1, where the total number of knots for the $year$ smoother, $s_2^*()$, was increased to 12 times the total number of years in the time series.  This approach potentially allows the smoother for $year$ to be unconstrained in how much variance is fit within and between years.

@@

* start with description of model performance, which models had the best performance (gam0 v gam6, boxcox v log)
* then describe figure 4

@@

# use previously fitted list of models
trans <- 'boxcox'
mods <- list(
  gam0 = anlz_gam(tomod, mod = 'gam0', trans = trans),
  gam1 = anlz_gam(tomod, mod = 'gam1', trans = trans),
  gam2 = anlz_gam(tomod, mod = 'gam2', trans = trans),
  gam6 = anlz_gam(tomod, mod = 'gam6', trans = trans)
)

p1 <- show_prddoy(mods = mods, ylab = 'Chlorophyll-a (ug/L)', nfac = 2, size = 0.8, alpha = 0.8) + 
  theme(legend.position = 'right') + 
  guides(colour = ggplot2::guide_colourbar(barheight = 15, barwidth = 1)) +
  scale_colour_gradient(low = "white", high = "black") + 
  theme_minimal() + 
  theme(
    legend.title = element_blank()
  )

png('figs/modsumfig.png', height = 4.5, width = 6, units = 'in', family = 'serif', res = 400)
print(p1)
dev.off()
```
```{r modsumfig, fig.cap = "General output for each of the four GAM structures in equations \\@ref(eq:gamzero) to \\@ref(eq:gamsix).  Day of year (Julian) is on the x-axis and each curve represents a different fitted year for chlorophyll concentration."}
knitr::include_graphics('figs/modsumfig.png')
```

```{r, results = 'hide', message = F}
# get predictions to plot
prdplo <- bind_rows(mods_chl18, mods_chl24, mods_chl27, mods_chl32) %>% 
  mutate(prddat = pmap(list(data, modi), function(data, modi){

@@

```{r modsumtab}
tab <- tibble(
  GAM = c('`gam0`', '`gam1`', '`gam2`', '`gam6`'),
  Summary = c(
    'Linear trend with seasonality', 
    'Nonlinear trends seasonality (constrained knots)', 
    'Nonlinear trends wiht seasonality, plus interaction',
    'Nonlinear trends with seasonality (unconstrained knots)'
  ),
  Details = c(
    'A linear trend in time over the course of the time series (year); and Within-year seasonal fluctuations that follow the same pattern every year, e.g., large spring bloom and small fall bloom (`s(doy, bs = "cc")`)',
    'Treats year and doy same as `gam0`; and A smooth non-linear trend through time, such that the seasonal peaks are similar across years, but interannual change is evident', 
    'Treats cyear and doy same as `gam1`; and Within-year seasonal fluctuations that vary across years but in a constrained pattern compared to `gam6` (`ti(year, doy, bs = c("tp", "cc"))`)', 
    'Variant of `gam1` with much higher values of k, allowing the splines to follow greater fluctuations both within and across years.')
  )

@@

```{r modprftab, tab.cap = 'Model performance statistics for each station and GAM structure (Table \\@ref(tab:modsumtab)). Results are for models of chlorophyll-a using BoxCox and logarathmic transformations. Four GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `gam0`),  $\\Delta$ GCV, and R-squared values.  Probability values ("P-val") show the results of Analysis of Variance comparisons between all BoxCox or logarithmic models at a station.  The p-value for a row refers to the comparison of the model in the current row to the one preceeding (`gam0` is the base model and does not have a p-value).  A significant p-value indicates the model in the current row provides a statistically significant difference in explaining chlorophyll variance from the model in the preceding row.  ns: not significant at $\\alpha = 0.05$, * $p< 0.05$, and ** $p < 0.005$.'}

>>>>

@@

A challenge for optimizing GAMs to time series data with significant intra- and interannual variation is choosing an appropriate functional form (i.e., gam1, gam2, or gam6) and determining how much variation could be explained by the smoothers in each function.  The four models above provide tradeoffs in the functional forms that balance descriptive and computational efficiency (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).  However, the individual smoothing functions (i.e, $s_1()$, $s_2()$, $s_2^*()$ and $ti_1()$) require user input on the potential upper limit for the *k*-values that could be used during model fitting.  Initial testing by Murphy et al. (2019) showed that the default *k* values in the mgcv package were insufficient for describing the interannual variation in the response variable.  Instead, the upper limit for the number of knots for the $s_2(year)$ smoother was specified as the maximum between 10 or  $2 / 3 \times$ the number of years in the time series.   We follow the same approach herein for gam1 and gam2.  Further, the default *k*-values for $s_2()$ and $ti_1()$ were considered appropriate and not tested further. For the application to SFE chlorophyll data, gam6 was also added as an extension of gam1, where the total number of knots for the $year$ smoother, $s_2^*()$, was increased to 12 times the total number of years in the time series.  This approach potentially allows the smoother for $year$ to be unconstrained in how much variance is fit within and between years.

@@

* start with description of model performance, which models had the best performance (gam1 v gam6, boxcox v log)
* then describe figure 4

@@

# use previously fitted list of models
trans <- 'boxcox'
mods <- list(
  gam1 = anlz_gam(tomod, mod = 'gam1', trans = trans),
  gam2 = anlz_gam(tomod, mod = 'gam2', trans = trans),
  gam6 = anlz_gam(tomod, mod = 'gam6', trans = trans)
)

p1 <- show_prddoy(mods = mods, ylab = 'Chlorophyll-a (ug/L)', nfac = 3, size = 0.8, alpha = 0.8) + 
  theme(legend.position = 'right') + 
  guides(colour = ggplot2::guide_colourbar(barheight = 15, barwidth = 1)) +
  scale_colour_gradient(low = "white", high = "black") + 
  theme_minimal() + 
  theme(
    legend.title = element_blank()
  )

png('figs/modsumfig.png', height = 2.5, width = 6, units = 'in', family = 'serif', res = 400)
print(p1)
dev.off()
```
```{r modsumfig, fig.cap = "General output for each of the four GAM structures in equations \\@ref(eq:gamzero) to \\@ref(eq:gamsix).  Day of year (Julian) is on the x-axis and each curve represents a different fitted year for chlorophyll concentration."}
knitr::include_graphics('figs/modsumfig.png')
```

```{r, results = 'hide', message = F}
# get predictions to plot
prdplo <- bind_rows(mods_chl18, mods_chl24, mods_chl27, mods_chl32) %>% 
  filter(model != 'gam0') %>% 
  mutate(prddat = pmap(list(data, modi), function(data, modi){

@@

```{r modsumtab}
tab <- tibble(
  GAM = c('`gam1`', '`gam2`', '`gam6`'),
  Summary = c(
    'Nonlinear trends seasonality (constrained knots)', 
    'Nonlinear trends wiht seasonality, plus interaction',
    'Nonlinear trends with seasonality (unconstrained knots)'
  ),
  Details = c(
    'Treats year and doy with separate smoothers with a smooth non-linear trend through time, such that the seasonal peaks are similar across years, but interannual change is evident', 
    'Treats cyear and doy same as `gam1`; and Within-year seasonal fluctuations that vary across years but in a constrained pattern compared to `gam6` (`ti(year, doy, bs = c("tp", "cc"))`)', 
    'Variant of `gam1` with much higher values of k, allowing the splines to follow greater fluctuations both within and across years.')
  )

@@

```{r modprftab, tab.cap = 'Model performance statistics for each station and GAM structure (Table \\@ref(tab:modsumtab)). Results are for models of chlorophyll-a using BoxCox and logarathmic transformations. Three GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `gam1`),  $\\Delta$ GCV, and R-squared values.  Probability values ("P-val") show the results of Analysis of Variance comparisons between all BoxCox or logarithmic models at a station.  The p-value for a row refers to the comparison of the model in the current row to the one preceeding (`gam1` is the base model and does not have a p-value).  A significant p-value indicates the model in the current row provides a statistically significant difference in explaining chlorophyll variance from the model in the preceding row.  ns: not significant at $\\alpha = 0.05$, * $p< 0.05$, and ** $p < 0.005$.'}
